### 2015-10-26 pde/pdetest1.tex
added laplace

\documentclass{article}
\usepackage{amssymb,amsmath}
\usepackage{mathtools}
\usepackage{graphicx}
\topmargin=-0.3in
\textheight=9.2in
\textwidth=168mm
\oddsidemargin=-0.2in
\evensidemargin=-0.2in
\begin{document}
%\indent Let's look at another second order PDE: the wave equation. This equation will be a little harder to solve due to the fact that it contains two second order derivates. Let's take a look:\\
%\noindent\textbf{\textit{Ex:}} Solve the following PDE
%\[
%\begin{cases*}
%u_{tt} - u_{xx} = 0\\
%u(0,t)=u(L,t) = 0\\
%u(x,0) = f(x),\ u_{t}(x,0) = g(x)
%\end{cases*}
%\]
%\indent Let's take a second before we jump into solving this equation to analyze it. We have a homogeneous, second order PDE. We also have boundary conditions, much like with the heat equation. The initial conditions are a little strange, but we won't get into them until we discuss Fourier series.\\
%\noindent We will use separation of variables to solve this equation. With this technique, we know that we are looking for a solution in the form of 
%\[
%u(x,t) = \xi(x)\phi(t).
%\]
%\noindent So, we substitute into the original PDE, and we get as follows:
%\begin{gather*}
%u_{tt} - u_{xx} = 0 =\phi^{''}\xi - \xi^{''}\phi\\
%\frac{\phi^{''}}{\phi} = \frac{\xi^{''}}{\xi} = \lambda
%\end{gather*}
%\noindent Now we need to find our eigenvalues and eigenfunctions. There's a trick here, however. In the heat equation, we had the liberty of solving the time equation whenever we wanted to. Here, we don't have that ability just yet, since $\phi(t)$ is ``trapped'' in a second order ODE. The only real choice we have here is to do what we did earlier, and apply our boundary conditions to the $\xi(x)$ ODE. We won't delve into the process for finding the eigenvalues, so here's the only non-trivial pair:
%\[
%\{\lambda_{n} = (n\pi)^{2},\ \sin{(n\pi x)}\}
%\]
%\indent Now let's take a second to imagine that we had had boundary conditions for our second order ODE. We would have had the same setup for finding the eigenvalues, but our eigenfunction would have involved $t$ instead of $x$.

\newpage
\section{First Order Partial Differential Equations}
\hrule
\noindent\\
\noindent\textbf{\textit{Ex:}} Solve the following PDE
\[
\begin{cases*}
yu_{x} + xu_{y} = 0\\
u(0,y) = e^{-y^{2}}
\end{cases*}
\]
\indent \textbf{\textit{Solution:}} This is a nice PDE; there are two ways to solve it, but we will focus on the easier one. We can see that we have a coefficient on both $u_{x}$ and $u_{y}$. This tells us that when $u$ was differentiated, there was a constant $y$ and $x$. So now we can say that
\[
\frac{dy}{dx} = \frac{x}{y}
\]
\noindent This is a simple ODE to solve. We can use separation of variables:
\begin{align*}
xdx &= ydy\\
\int{x dx} &= \int{y dy}\\
\frac{1}{2}x^{2} &= \frac{1}{2}y^{2} + c\\
x^{2} - y^{2} &= c_{0}\\
y^{2} - x^{2} &= c_{1}
\end{align*}
\noindent You'll notice that we multiplied both sides by $-1$. This is for convenience. Now we can express the answer as
\[
u(x,y) = f(y^{2} - x^{2})
\]
\noindent And we are done.

@@

\noindent \textbf{\textit{Ex:}} Solve the following PDE
\[
\begin{cases*}
u_{tt} - u_{xx} = 1 + t\cos{(2\pi x)}\\
u_{x}(0,t) = u_{x}(1,t) = 0\\
u(x,0) = 2\cos{(2\pi x)},\quad u_{t}(x,0) = 1 + \cos{(\pi x)} - \cos{(2\pi x)}
\end{cases*}
\]
\indent \textbf{\textit{Solution:}} This problem is an example of the wave equation with a non-homogeneous solution. We are lucky this time; the boundary conditions are already homogeneous, so there is no need to change our variable. We know that since both of the boundary conditions are derivatives, we are dealing with a Fourier cosine series. This yields the eigenvalue/function of
\[
\left\{\lambda_{n} = (n\pi)^{2}, \cos{(n\pi x)}\right\}
\]
Since we are dealing with a cosine series, we also have a leading term, $a_{0}$. Our full series looks like
\[
u(x,t) = a_{0}(t) + \sum_{n=1}^{\infty} a_{n}(t)\cos{(n\pi x)}
\]
We need $u_{tt}$ and $u_{xx}$, so we have:
\[
u_{tt}(x,t) = a_{0}''(t) + \sum_{n=1}^{\infty} a_{n}''(t)\cos{(n\pi x)} 
\qquad \text{and} \qquad 
u_{xx}(x,t) = -\sum_{n=1}^{\infty} (n\pi)^{2}a_{n}(t)\cos{(n\pi x)}
\]
All we have to do now is simply plug into our original equation. This gives us
\begin{align*}
a_{0}''(t) + \sum_{n=1}^{\infty} a_{n}''(t)\cos{(n\pi x)} + \sum_{n=1}^{\infty} (n\pi)^{2}a_{n}(t)\cos{(n\pi x)} &= 1 + t\cos{(2\pi x)}\\
a_{0}''(t) + \sum_{n=1}^{\infty}(a_{n}''(t) + (n\pi)^{2}a_{n}(t))\cos{(n\pi x)} &= 1 + t\cos{(2\pi x)}
\end{align*}
Now we can play the matching game. We have a $constant + series\ term$ on the right side, and $constant + series$ on the left, so we can match up the constants. This gives us 
\[
a_{0}''(t) = 1
\]
Now we match up the series term. On the right hand side, we have $\cos{(2\pi x)}$, so we know that for $n= 2$, we have
\[
a_{2}''(t) + 4\pi^{2}a_{2}(t) = t
\]
$a_{0}$ and $a_{2}$ are the only terms that appear, so we can assume for now that $\forall n \neq 2\ \text{and}\ 3, a_{n} = 0$. We are left with two ODE's, so we now need to find the corresponding initial conditions. Let's apply the first initial condition to our PDE. This gives us
\[a_{0}(0) + \sum_{n=1}^{\infty}a_{n}(0)\cos{(n\pi x)} = 2\cos{(2\pi t)}\]
Which tells us
\[a_{0} = 0 \qquad\qquad \qquad\qquad a_{2} = 2\]
Now we can apply the second initial condition. After taking the partial derivative of $u$ with respect to $t$, we have
\[
a_{0}'(0) + \sum_{n=1}^{\infty}a_{n}'(0)\cos{(n\pi x)} = 1 + \cos{(\pi x)} - \cos{(2\pi x)}
\]
Some more clever pattern matching give us
\[
a_{0}'(0) = 1 \qquad\qquad a_{1}'(0) = 1 \qquad\qquad a_{2}'(0) = -1
\]
Notice that we have a term $a_{1}'$. We cannot simply ignore this value; rather we have to take it into account when we find our solution. We know that we have an ODE for $a_{1}(t)$, but it may not be evident how to express it. Recall that earlier we said $\forall n \neq 2\ \text{and}\ 3, a_{n} = 0$. This means that $a_{1} = 0$, and since $a_{1}$ is a coefficient in the Fourier series, we know that $a_{1}''(t) + \pi^{2}a_{1}(t) = 0$. We now have three different ODE's that we must solve:
\begin{alignat*}{3}
&\begin{cases}
a_{0}''(t) = 1\\
a_{0}'(0) = 1,\quad a_{0}(0) = 1
\end{cases}
&\begin{cases}
a_{1}''(t) + \pi^{2}a_{1}(t) = 0\\
a_{1}'(0) = 1,\quad a_{1}(0) = 0
\end{cases}
&&\begin{cases}
a_{2}''(t) + 4\pi^{2}a_{2}(t) = t\\
a_{2}'(0) = 1,\quad a_{2}(0) = 2
\end{cases}
\end{alignat*}
Luckily, Mathematica is here to save the day. We will skip the work and let the computer do the thinking for us here. We get that
\begin{align*}
a_{0}(t) &= \frac{1}{2}t^{2} + t\\
a_{1}(t) &= \frac{\pi\cos{(t\pi)} + \sin{(t\pi)}}{\pi}\\
a_{2}(t) &= \frac{2\pi t + 16\pi^{3}\cos{(2\pi t)} - \sin{(2\pi t)} + 4\pi^{2}\sin{(2\pi t)}}{8\pi^{3}}
\end{align*}
Now all we need to do is plug back in to our $u$. Our final answer becomes
\[
u(x,t) = \frac{1}{2}t^{2} + t + \frac{\pi\cos{(t\pi)} + \sin{(t\pi)}}{\pi}\cos{(\pi x)} +  \frac{2\pi t + 16\pi^{3}\cos{(2\pi t)} - \sin{(2\pi t)} + 4\pi^{2}\sin{(2\pi t)}}{8\pi^{3}}\cos{(2\pi x)}
\]
And we are done.
\newpage

\subsection{The Laplace Equation}
\noindent \\
\indent Up to this point, we have only looked at the wave equation and the heat equation. There is, however, one other equation in the classical trinity: the Laplace Equation. We haven't dealt with this equation yet because we have to be very careful in our approach to solving it. When we looked at the homogeneous wave and heat equations, we used the general approach called \textit{Separation of Variables}. In the case of the homogeneous Laplace equation, we may not be able to tackle the problem using separation of variables. The problem arises from the various types of boundaries that effect the domain. We can either have a 'rectangular' bound, which will be separable, or a 'circular' bound which is \textbf{not} separable. Imagine that we have the following two problems:
\begin{align*}
&A =\begin{cases}
u_{xx} + u_{yy} = 0\quad \text{on}\ [0,1]\times[0,2]\\
u(x,0) = u(x,2) = 0\\
u(0,y) = y, u(1,y) = 0
\end{cases}
&B =\begin{cases}
u_{xx} + u_{yy} = 0\quad \text{on}\ x^{2} + y^{2} < a^{2}\\
u(x,y)\ \big |_{x^{2} + y^{2} = a^{2}} = f(x,y)
\end{cases}
\end{align*}
$A$ is an example of a separable Laplace equation. We can see graphically that the domain $[0,1] \times [0,2]$ is rectangular. $B$'s domain is not separable, since it is circular. Let's take a look $A$.\\\\
\noindent \textbf{\textit{Ex:}} Solve the following PDE
\[
\begin{cases*}
u_{xx} + u_{yy} = 0\quad \text{on}\ [0,1]\times[0,2]\\
u(x,0) = u(x,2) = 0\\
u(0,y) = y, u(1,y) = 0
\end{cases*}
\]
\indent \textbf{\textit{Solution:}} Earlier, we discussed the different boundary conditions we can encounter and their corresponding eigenvalues/functions. In this problem, we have Dirichlet boundary conditions. This tells us that
\[
\left\{\lambda_{n} = \left(\frac{n\pi}{2}\right)^{2}, \phi_{n}(y) = \sin{\left(\frac{n\pi y}{2}\right)}\right\}
\]
Notice that we used $y$ here instead of $x$. We did this because the boundary conditions are defined in terms of $y$, not $x$. Now we can continue to solve our problem. Recall that our separation of variables gives us two second order ODE's,
\[
X'' - \lambda X = 0 \qquad\qquad \text{and} \qquad\qquad \lambda Y - Y'' = 0
\]
We already solved for $Y(y)$ by finding the corresponding eigenvalues/vectors. Now we plug in our $\lambda_{n}$ into our $X$ ODE. Solving the resulting ODE will give us a basis of
\[
\left\{ e^{\frac{n\pi}{2}x}, e^{-\frac{n\pi}{2}x} \right\}
\]
Now we could continue at this point, but if we think ahead, we can see that when we eventually apply our initial conditions, nothing will cancel out. At this point, we want to transform our basis. We have as follows
\[
\left\{ e^{\frac{n\pi}{2}x}, e^{-\frac{n\pi}{2}x} \right\} \implies \left\{ \cosh{\left(\frac{n\pi x}{2}\right)}, \sinh{\left(\frac{n\pi x}{2}\right)} \right\}
\]
All we did here is apply an identity. Now we have everything we need to set up a solution. Our $u(x,y)$ will be
\[
u(x,y) = \sum_{n=1}^{\infty}\left[a_{n}\cosh{\left(\frac{n\pi x}{2}\right)} + b_{n}\sinh{\left(\frac{n\pi x}{2}\right)}\right]\sin{\left(\frac{n\pi y}{2}\right)}
\]
We can apply our initial conditions, which gives us $b_{n}\ \text{and}\ a_{n}$:
\begin{align*}
a_{n} &= \frac{4(-1)^{n+1}}{n\pi}\\
b_{n} &= -\coth{\left(\frac{n \pi}{2}\right)}\frac{4(-1)^{n+1}}{n\pi}
\end{align*}
This yields a final solution of
\[
u(x,y) = \sum_{n=1}^{\infty}\left[ \frac{4(-1)^{n+1}}{n\pi} \cosh{\left(\frac{n\pi x}{2}\right)} -\coth{\left(\frac{n \pi}{2}\right)}\frac{4(-1)^{n+1}}{n\pi}\sinh{\left(\frac{n\pi x}{2}\right)}\right]\sin{\left(\frac{n\pi y}{2}\right)}
\]
And we are done
\newpage

\indent Now we turn our attention to the Laplace Equation with a circular domain. As we discussed earlier, we can't simply say that our solution will be in the form of $X(x)Y(y)$. However, we can force the domain to be rectangular if we use polar coordinates. The translation is not too difficult. Let's consider $B$ from above.\\\\

\noindent \textbf{\textit{Ex:}} Solve the following PDE
\[
\begin{cases*}
u_{xx} + u_{yy} = 0\quad \text{on}\ x^{2} + y^{2} < a^{2}\\
u(x,y)\ \big |_{x^{2} + y^{2} = a^{2}} = f(x,y)
\end{cases*}
\]
\indent \textbf{\textit{Solution:}} We want to express the partial derivatives here in terms of $r\ \text{and}\ \theta$. Recall that we have
\[
x = r\cos{\theta} \qquad\qquad \text{and} \qquad\qquad y = r\sin{\theta}
\]
First, we will find the partial derivatives of $x$ and $y$ with respect to $r$ and $\theta$. Since $x = r\cos{\theta}$, 
\[
\frac{\partial x}{\partial r} = \cos{\theta} \qquad\qquad \frac{\partial x}{\partial \theta} = -r\sin{\theta}
\]
We can compute the derivatives for $y$ in a similar fashion:
\[
\frac{\partial y}{\partial r} = \sin{\theta} \qquad\qquad \frac{\partial y}{\partial \theta} = r\cos{\theta}
\]
Now we need to find $u_{rr}$ and $u_{\theta\theta}$. Let's focus on $u_{rr}$ first. We will need to start by finding $u_{r}$. We have as follows
\begin{align*}
\frac{\partial u}{\partial r}& = \frac{\partial u}{\partial x}\frac{\partial x}{\partial r} + \frac{\partial u}{\partial y}\frac{\partial y}{\partial r}\\
&= \cos{\theta}\frac{\partial u}{\partial x} + \sin{\theta}\frac{\partial u}{\partial y}
\end{align*}
We have $u_{r}$, so now we need to compute $u_{rr}$. We have as follows:
\begin{align*}
\frac{\partial^{2} u}{\partial r^{2}} &= \cos{\theta}\frac{\partial }{\partial r}\frac{\partial u}{\partial x} + \sin{\theta}\frac{\partial }{\partial r}\frac{\partial u}{\partial y}\\
&= \cos^{2}{\theta}\frac{\partial^{2} u}{\partial x^{2}} + 2\cos{\theta}\sin{\theta}\frac{\partial^{2}u}{\partial x\partial y} + \sin^{2}{\theta}\frac{\partial^{2} u}{\partial y^{2}}
\end{align*}
We won't delve into finding $u_{\theta\theta}$, as the partials get a little out of hand. However, we end up with
\begin{align*}
\frac{\partial^{2}u}{\partial r^{2}} + \frac{1}{r^{2}}\frac{\partial^{2}u}{\partial\theta^{2}} &= -\frac{1}{r}\frac{\partial u}{\partial r} + \frac{\partial^{2} u}{\partial x^{2}} + \frac{\partial^{2} u}{\partial y^{2}}\\
u_{xx} + u_{yy} &= u_{rr} + \frac{1}{r}u_{r} + \frac{1}{r^{2}}u_{\theta\theta} = 0
\end{align*}
We now have the Laplace equation represented in the polar coordinate system. Note that the translation of the boundary condition is simple. It becomes a function of only $\theta$. We have the following system:
\[
\begin{cases*}
u_{rr} + \frac{1}{r}u_{r} + \frac{1}{r^{2}}u_{\theta\theta} = 0\quad \text{on}\ 0 < r < a\\
u(\theta) = f(\theta)\quad \text{on}\ 0 < \theta < 2\pi
\end{cases*}
\]
Now we can use separation of variables to solve the PDE. The steps are the same (although this time we will use $R(r)\ \text{and}\ T(\theta)$). We will end up with the following equivalence:
\[
\frac{r^{2}R'' + rR'}{R} = \frac{T''}{T} = \lambda
\]
Since we have a Dirichlet boundary conditions, we know that
\[
\lambda_{n} = \frac{n^{2}}{4}
\]
Plugging this into our equation for $R$ gives us
\[
r^{2}R'' + rR' - \frac{n^{2}}{4}R = 0 
\]
\end{document}

>>>>

\documentclass{article}
\usepackage{amssymb,amsmath}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{multicol}
\topmargin=-0.3in
\textheight=9.2in
\textwidth=168mm
\oddsidemargin=-0.2in
\evensidemargin=-0.2in
\begin{document}
\section{First Order Partial Differential Equations}
\hrule
\noindent\\
\noindent\textbf{\textit{Ex:}} Solve the following PDE
\[
\begin{cases*}
yu_{x} + xu_{y} = 0\\
u(0,y) = e^{-y^{2}}
\end{cases*}
\]
\indent \textbf{\textit{Solution:}} This is a nice PDE; there are two ways to solve it, but we will focus on the easier one. We can see that we have a coefficient on both $u_{x}$ and $u_{y}$. This tells us that when $u$ was differentiated, there was a constant $y$ and $x$. So now we can say that
\[
\frac{dy}{dx} = \frac{x}{y}
\]
\noindent This is a simple ODE to solve. We can use separation of variables:
\begin{align*}
xdx &= ydy\\
\int{x dx} &= \int{y dy}\\
\frac{1}{2}x^{2} &= \frac{1}{2}y^{2} + c\\
x^{2} - y^{2} &= c_{0}\\
y^{2} - x^{2} &= c_{1}
\end{align*}
\noindent You'll notice that we multiplied both sides by $-1$. This is for convenience. Now we can express the answer as
\[
u(x,y) = f(y^{2} - x^{2})
\]
\noindent And we are done.

@@

u_{tt} - u_{xx} = 1 + t\cos{(2\pi x)}\\
u_{x}(0,t) = u_{x}(1,t) = 0\\
u(x,0) = 2\cos{(2\pi x)},\quad u_{t}(x,0) = 1 + \cos{(\pi x)} - \cos{(2\pi x)}
\indent \textbf{\textit{Solution:}} This problem is an example of the wave equation with a non-homogeneous solution. We are lucky this time; the boundary conditions are already homogeneous, so there is no need to change our variable. We know that since both of the boundary conditions are derivatives, we are dealing with a Fourier cosine series. This yields the eigenvalue/function of
\left\{\lambda_{n} = (n\pi)^{2}, \cos{(n\pi x)}\right\}
Since we are dealing with a cosine series, we also have a leading term, $a_{0}$. Our full series looks like
u(x,t) = a_{0}(t) + \sum_{n=1}^{\infty} a_{n}(t)\cos{(n\pi x)}
We need $u_{tt}$ and $u_{xx}$, so we have:
u_{tt}(x,t) = a_{0}''(t) + \sum_{n=1}^{\infty} a_{n}''(t)\cos{(n\pi x)} 
\qquad \text{and} \qquad 
u_{xx}(x,t) = -\sum_{n=1}^{\infty} (n\pi)^{2}a_{n}(t)\cos{(n\pi x)}
All we have to do now is simply plug into our original equation. This gives us
a_{0}''(t) + \sum_{n=1}^{\infty} a_{n}''(t)\cos{(n\pi x)} + \sum_{n=1}^{\infty} (n\pi)^{2}a_{n}(t)\cos{(n\pi x)} &= 1 + t\cos{(2\pi x)}\\
a_{0}''(t) + \sum_{n=1}^{\infty}(a_{n}''(t) + (n\pi)^{2}a_{n}(t))\cos{(n\pi x)} &= 1 + t\cos{(2\pi x)}
Now we can play the matching game. We have a $constant + series\ term$ on the right side, and $constant + series$ on the left, so we can match up the constants. This gives us 
a_{0}''(t) = 1
Now we match up the series term. On the right hand side, we have $\cos{(2\pi x)}$, so we know that for $n= 2$, we have
a_{2}''(t) + 4\pi^{2}a_{2}(t) = t
$a_{0}$ and $a_{2}$ are the only terms that appear, so we can assume for now that $\forall n \neq 2\ \text{and}\ 3, a_{n} = 0$. We are left with two ODE's, so we now need to find the corresponding initial conditions. Let's apply the first initial condition to our PDE. This gives us
\[a_{0}(0) + \sum_{n=1}^{\infty}a_{n}(0)\cos{(n\pi x)} = 2\cos{(2\pi t)}\]
Which tells us
\[a_{0} = 0 \qquad\qquad \qquad\qquad a_{2} = 2\]
Now we can apply the second initial condition. After taking the partial derivative of $u$ with respect to $t$, we have
a_{0}'(0) + \sum_{n=1}^{\infty}a_{n}'(0)\cos{(n\pi x)} = 1 + \cos{(\pi x)} - \cos{(2\pi x)}
Some more clever pattern matching give us
a_{0}'(0) = 1 \qquad\qquad a_{1}'(0) = 1 \qquad\qquad a_{2}'(0) = -1
Notice that we have a term $a_{1}'$. We cannot simply ignore this value; rather we have to take it into account when we find our solution. We know that we have an ODE for $a_{1}(t)$, but it may not be evident how to express it. Recall that earlier we said $\forall n \neq 2\ \text{and}\ 3, a_{n} = 0$. This means that $a_{1} = 0$, and since $a_{1}$ is a coefficient in the Fourier series, we know that $a_{1}''(t) + \pi^{2}a_{1}(t) = 0$. We now have three different ODE's that we must solve:
\begin{alignat*}{3}
&\begin{cases}
a_{0}''(t) = 1\\
a_{0}'(0) = 1,\quad a_{0}(0) = 1
&\begin{cases}
a_{1}''(t) + \pi^{2}a_{1}(t) = 0\\
a_{1}'(0) = 1,\quad a_{1}(0) = 0
&&\begin{cases}
a_{2}''(t) + 4\pi^{2}a_{2}(t) = t\\
a_{2}'(0) = 1,\quad a_{2}(0) = 2
\end{alignat*}
Luckily, Mathematica is here to save the day. We will skip the work and let the computer do the thinking for us. We get that
a_{0}(t) &= \frac{1}{2}t^{2} + t\\
a_{1}(t) &= \frac{\pi\cos{(t\pi)} + \sin{(t\pi)}}{\pi}\\
a_{2}(t) &= \frac{2\pi t + 16\pi^{3}\cos{(2\pi t)} - \sin{(2\pi t)} + 4\pi^{2}\sin{(2\pi t)}}{8\pi^{3}}
Now all we need to do is plug back in to our $u$. Our final answer becomes
u(x,t) = \frac{1}{2}t^{2} + t + \frac{\pi\cos{(t\pi)} + \sin{(t\pi)}}{\pi}\cos{(\pi x)} +  \frac{2\pi t + 16\pi^{3}\cos{(2\pi t)} - \sin{(2\pi t)} + 4\pi^{2}\sin{(2\pi t)}}{8\pi^{3}}\cos{(2\pi x)}
Just for interest, the plot of this solution looks like
\begin{center}
\includegraphics[scale=0.6]{pg23_plot} 
\end{center}
On $-\pi \leq \theta \leq \pi$ and $-\pi \leq x \leq \pi$; and we are done.

\subsection{The Laplace Equation}
\noindent \\
\indent Up to this point, we have only looked at the wave equation and the heat equation. There is, however, one other equation in the classical trinity: the Laplace Equation. We haven't dealt with this equation yet because we have to be very careful in our approach to solving it. When we looked at the homogeneous wave and heat equations, we used the general approach called \textit{Separation of Variables}. In the case of the homogeneous Laplace equation, we may not be able to tackle the problem using separation of variables. The problem arises from the various types of boundaries that effect the domain. We can either have a 'rectangular' bound, which will be separable, or a 'circular' bound which is \textbf{not} separable. Imagine that we have the following two problems:
\begin{align*}
&A =\begin{cases}
u_{xx} + u_{yy} = 0\quad \text{on}\ [0,1]\times[0,2]\\
u(x,0) = u(x,2) = 0\\
u(0,y) = y, u(1,y) = 0
\end{cases}
&B =\begin{cases}
u_{xx} + u_{yy} = 0\quad \text{on}\ x^{2} + y^{2} < a^{2}\\
u(x,y)\ \big |_{x^{2} + y^{2} = a^{2}} = f(x,y)
\end{cases}
\end{align*}
$A$ is an example of a separable Laplace equation. We can see graphically that the domain $[0,1] \times [0,2]$ is rectangular. $B$'s domain is not separable, since it is circular. Let's take a look at $A$.\\\\
\noindent \textbf{\textit{Ex:}} Solve the following PDE
\[
\begin{cases*}
u_{xx} + u_{yy} = 0\quad \text{on}\ [0,1]\times[0,2]\\
u(x,0) = u(x,2) = 0\\
u(0,y) = y, u(1,y) = 0
\end{cases*}
\]
\indent \textbf{\textit{Solution:}} Earlier, we discussed the different boundary conditions we can encounter and their corresponding eigenvalues/functions. In this problem, we have Dirichlet boundary conditions. This tells us that
\[
\left\{\lambda_{n} = \left(\frac{n\pi}{2}\right)^{2}, \phi_{n}(y) = \sin{\left(\frac{n\pi y}{2}\right)}\right\}
\]
Notice that we used $y$ here instead of $x$. We did this because the boundary conditions are defined in terms of $y$, not $x$. Now we can continue to solve our problem. Recall that our separation of variables gives us two second order ODE's,
\[
X'' - \lambda X = 0 \qquad\qquad \text{and} \qquad\qquad \lambda Y - Y'' = 0
\]
We already solved for $Y(y)$ by finding the corresponding eigenvalues/vectors. Now we plug in our $\lambda_{n}$ into our $X$ ODE. Solving the resulting ODE will give us a basis of
\[
\left\{ e^{\frac{n\pi}{2}x}, e^{-\frac{n\pi}{2}x} \right\}
\]
Now we could continue at this point, but if we think ahead, we can see that when we eventually apply our initial conditions, nothing will cancel out. At this point, we want to transform our basis. We have as follows
\[
\left\{ e^{\frac{n\pi}{2}x}, e^{-\frac{n\pi}{2}x} \right\} \implies \left\{ \cosh{\left(\frac{n\pi x}{2}\right)}, \sinh{\left(\frac{n\pi x}{2}\right)} \right\}
\]
All we did here is apply an identity. Now we have everything we need to set up a solution. Our $u(x,y)$ will be
\[
u(x,y) = \sum_{n=1}^{\infty}\left[a_{n}\cosh{\left(\frac{n\pi x}{2}\right)} + b_{n}\sinh{\left(\frac{n\pi x}{2}\right)}\right]\sin{\left(\frac{n\pi y}{2}\right)}
\]
We can apply our initial conditions, which gives us $b_{n}\ \text{and}\ a_{n}$:
\begin{align*}
a_{n} &= \frac{4(-1)^{n+1}}{n\pi}\\
b_{n} &= -\coth{\left(\frac{n \pi}{2}\right)}\frac{4(-1)^{n+1}}{n\pi}
\end{align*}
This yields a final solution of
\[
u(x,y) = \sum_{n=1}^{\infty}\left[ \frac{4(-1)^{n+1}}{n\pi} \cosh{\left(\frac{n\pi x}{2}\right)} -\coth{\left(\frac{n \pi}{2}\right)}\frac{4(-1)^{n+1}}{n\pi}\sinh{\left(\frac{n\pi x}{2}\right)}\right]\sin{\left(\frac{n\pi y}{2}\right)}
\]
And we are done
\newpage

\indent Now we turn our attention to the Laplace Equation with a circular domain. As we discussed earlier, we can't simply say that our solution will be in the form of $X(x)Y(y)$. However, we can force the domain to be rectangular if we use polar coordinates. Let's consider a variation of $B$ from above.\\\\
u_{xx} + u_{yy} = 0\quad \text{on}\ x^{2} + y^{2} < a^{2}\\
u(a,\theta) = U_{1}\quad \text{on}\ 0 \leq \theta \leq \pi\\
u(a,\theta) = U_{2}\quad \text{on}\ \pi \leq \theta \leq 2\pi
\indent \textbf{\textit{Solution:}} We want to express the partial derivatives here in terms of $r\ \text{and}\ \theta$. Recall that we have
x = r\cos{\theta} \qquad\qquad \text{and} \qquad\qquad y = r\sin{\theta}
First, we will find the partial derivatives of $x$ and $y$ with respect to $r$ and $\theta$. Since $x = r\cos{\theta}$, 
\frac{\partial x}{\partial r} = \cos{\theta} \qquad\qquad \frac{\partial x}{\partial \theta} = -r\sin{\theta}
We can compute the derivatives for $y$ in a similar fashion:
\frac{\partial y}{\partial r} = \sin{\theta} \qquad\qquad \frac{\partial y}{\partial \theta} = r\cos{\theta}
Now we need to find $u_{rr}$ and $u_{\theta\theta}$. Let's focus on $u_{rr}$ first. We will need to start by finding $u_{r}$. We have as follows
\frac{\partial u}{\partial r}& = \frac{\partial u}{\partial x}\frac{\partial x}{\partial r} + \frac{\partial u}{\partial y}\frac{\partial y}{\partial r}\\
&= \cos{\theta}\frac{\partial u}{\partial x} + \sin{\theta}\frac{\partial u}{\partial y}
We have $u_{r}$, so now we need to compute $u_{rr}$. We have as follows:
\frac{\partial^{2} u}{\partial r^{2}} &= \cos{\theta}\frac{\partial }{\partial r}\frac{\partial u}{\partial x} + \sin{\theta}\frac{\partial }{\partial r}\frac{\partial u}{\partial y}\\
&= \cos^{2}{\theta}\frac{\partial^{2} u}{\partial x^{2}} + 2\cos{\theta}\sin{\theta}\frac{\partial^{2}u}{\partial x\partial y} + \sin^{2}{\theta}\frac{\partial^{2} u}{\partial y^{2}}
We won't delve into finding $u_{\theta\theta}$, as the partials get a little out of hand. However, we end up with
\frac{\partial^{2}u}{\partial r^{2}} + \frac{1}{r^{2}}\frac{\partial^{2}u}{\partial\theta^{2}} &= -\frac{1}{r}\frac{\partial u}{\partial r} + \frac{\partial^{2} u}{\partial x^{2}} + \frac{\partial^{2} u}{\partial y^{2}}\\
u_{xx} + u_{yy} &= u_{rr} + \frac{1}{r}u_{r} + \frac{1}{r^{2}}u_{\theta\theta} = 0
Lets go ahead and formalize it:
\nabla^{2}u = u_{rr} + \frac{1}{r}u_{r} + \frac{1}{r^{2}}u_{\theta\theta}
Now we can treat this as a 2D problem; in other words, we can now use separation of variables to solve for $u$. As before, we know our solution will look like
u(r,\theta) = R(r)T(\theta)
Plugging everything in and realizing by the usual trick that everything s equal to a constant gives us
-\frac{T''}{T} = \frac{R'' + \frac{R'}{r}}{\frac{R}{r^{2}}} = \lambda
Since we are dealing in polar coordinates, we will go ahead and walk through the eigenvalues/functions. We will be solving for $T(\theta)$, since our boundary conditions are in terms of $\theta$. We also should note that the solutions will repeat themselves every $2\pi$, since we are in the polar system. In addition, our $u$ should be finite at $r=0$. Let's take a look at the eigenvalues for the Laplace equation.\\\\
\begin{minipage}[t]{0.32\textwidth}
\underline{$\lambda = 0$}\\
When $\lambda = 0$, we have \[T(\theta) = a\theta + b.\] This also tells us that \[R(r) = c\ln{(r)} + d.\] Putting everything together gives us \[ u(r,\theta) = (a\theta + b)(c\ln{r} + d).\] Since the solution is periodic on $2\pi$, we know that $a=0$. Since the solution has to remain finite at $r=0$, we know that $c = 0$. This gives us that $u = bd$, or that $u$ is equal to a constant.
\end{minipage}
\hspace{0.01\textwidth}
\begin{minipage}[t]{0.32\textwidth}
\underline{$\lambda > 0$}\\
With $\lambda > 0$, we have our characteristic equation $(m = \pm\ \sqrt{\lambda})$ yielding a real valued solution. This tells us that
T(\theta) = c_{1}\cosh{(\sqrt{\lambda}\theta)} + c_{2}\sinh{(\sqrt{\lambda}\theta)}.
The solution is periodic on $2\pi$, so we can say that $c_{1} = c_{2} = 0$. Therefore, we have a trivial solution.
\end{minipage}
\hspace{0.01\textwidth}
\begin{minipage}[t]{0.32\textwidth}
\underline{$\lambda < 0$}\\
$\lambda < 0$ gives us 
T(\theta) = c_{1}\cos{(\sqrt{\lambda}\theta)} + c_{2}\sin{(\sqrt{\lambda}\theta)}
This solution is periodic with $2\pi$, and it yields a trivial eigenvector.
\end{minipage}
\noindent\\\\\\
\noindent So we know that the only $\lambda$ that yields eigenvalues is $\lambda = 0$, and we know that this particular eigenvalue yields a constant solution. If we solve the differential equation for $R$, we will find that the only solution we care about is $r^{n}$. We can finally say that our solution will look like
u(r,\theta) = \frac{a_{0}}{2} + \sum_{n=1}^{\infty}r^{n}[a_{n}\cos{(n\theta)} + b_{n}\sin{(n\theta)}]
This solution is worth memorizing, as it will generally be the same for all variations of the Laplace equation similar to this one. Now we can apply the conditions for $\theta$. We have two conditions, but it is really just a piecewise function which we will call $f(\theta)$. We also know that our bound on the domain is $a$, or the length of the radius. This is enough information to find all the various variables in our $u(r,\theta)$. We have as follows:
u(a,\theta) = f(\theta) = \frac{a_{0}}{2} + \sum_{n=1}^{\infty}a^{n}[a_{n}\cos{(n\theta)} + b_{n}\sin{(n\theta)}]
Note that the $a$ which is the length of the radius is \textit{not} the same $a$ as $a_{n}$. Now we can solve for $a_{0}$ and $a_{n}$.
\begin{alignat*}{3}
a_{0} &= \frac{1}{\pi}\int_{0}^{2\pi}f(\theta)d\theta \qquad\qquad a_{n} &&= \frac{1}{\pi}\left[\int_{0}^{\pi}U_{1}\cos{(n\theta)}d\theta + \int_{\pi}^{2\pi}U_{2}\cos{(n\theta)}d\theta\right]\\
&= U_{1} + U_{2} &&=0
\end{alignat*}
For $b_{n}$, we have
a^{n}b_{n} &= \frac{1}{\pi}\left[\int_{0}^{\pi}U_{1}\sin{(n\theta)}d\theta + \int_{\pi}^{2\pi}U_{2}\sin{(n\theta)}d\theta\right]\\
&= \frac{1 - (-1)^{n}}{n\pi}(U_{1} - U_{2})\\
b_{n} &= \frac{\left[1 - (-1)^{n}\right](U_{1} - U_{2})}{n\pi a^{n}}\\
&= \frac{2(U_{1} - U_{2})}{(2k+1)\pi a^{2k+1}}
Putting it all together gives us a final solution of
